---
title: "lab3"
author: "Eleanore Brugh"
date: "2026-03-05"
output: html_document
---

#Lab3 Exercises
```{r packages}

library(spdep)
library(tidycensus)
library(sf)
library(tidyverse)
library(tmap)
library(stringr)

```

```{r}

options(tigris_use_cache = TRUE)

```


```{r data}

d.all <- get_acs(geography = "county",
                 table = "B15003",
                 output = "wide",
                 geometry = T,
                 year = 2020)

```

```{r}

glimpse(d.all)

```

```{r download variables}

df.vars <- load_variables(2020, "acs5", cache = T)

glimpse(df.vars)

```


```{r filter data to female population aged 75-79}

df.vars %>% dplyr::filter(concept == "SEX BY AGE") %>%
  dplyr::filter(str_detect(label, "Female")) %>%
  dplyr::filter(str_detect(label, "75"))

```


```{r a lot of data}

d.sex.by.age <- get_acs(geography = "county",
                        table = "B01001",
                        output = "wide",
                        geometry = T,
                        year = 2020)

glimpse(d.sex.by.age)

```


```{r}

ohio <- d.sex.by.age %>%
  mutate(., STATEID = stringr::str_sub(GEOID, 1, 2)) %>%
  dplyr::filter(STATEID == "39") 

tmap::tm_shape(ohio) + tm_polygons()

```

```{r checking coordinate system}

sf::st_crs(ohio)

```


```{r reproject}
#north america equidistant conic
ohio.projected <- ohio %>% sf::st_transform(., "ESRI:102010")

```

```{r map it again}

tmap::tm_shape(ohio.projected) + tm_polygons()

```

```{r making the neighborhood}

nb <- spdep::poly2nb(ohio.projected, queen = TRUE)

nb

#to see the neighbors in the first polygon in the object
nb[[1]]

```


```{r counties in index 1}
##Erie County Ohio
ohio.projected$NAME[1]

##Lorain, Sandusky, and Huron the neighbors
nb[[1]] %>% ohio.projected$NAME[.]

```

```{r assigning weights to each neighboring polygon}

#zero.policy=TRUE allows for lists of non-neighbors (use with caution)
lw <- spdep::nb2listw(nb, style="W", zero.policy=TRUE)

#see the weight of the first polygons four neighbors
lw$weights[1]

```

```{r}
#using the attribute to get the count of neighbors in W
neighbors <- attr(lw$weights,"comp")$d 

#plotting it
hist(neighbors)

```

```{r}
#average neighboring F 75-79 values
F75.lag <- lag.listw(lw, ohio.projected$B01001_047E)
F75.lag

```

###Morans 
```{r}

moran.test(ohio.projected$B01001_047E, lw)

```


###Moran's Plots
```{r}

moran.plot(ohio.projected$B01001_047E, lw, zero.policy=TRUE, plot=TRUE)

```



#MY TASKS
###Task 1
```{r poverty data}

options(tigris_use_cache = TRUE)

d.poverty <- get_acs(
  geography = "county",
  table = "B17001",
  output = "wide",
  geometry = TRUE,
  year = 2020
)

glimpse(d.poverty)

```

```{r subset}

#Ohio, Indian, Kentucky, West Virginia
four_states <- d.poverty %>%
  mutate(STATEID = stringr::str_sub(GEOID, 1, 2)) %>%
  filter(STATEID %in% c("39", "18", "21", "54"))

glimpse(four_states)
```

###Task 2
```{r normalize}
#the poverty counts and total population to determine the percent of people below the poverty line for each county
four_states <- four_states %>%
  mutate(poverty_pct = (B17001_002E / B17001_001E) * 100)

```

###Task 3
```{r histogram}

hist(
  four_states$poverty_pct,
  main = "County Poverty Rate For OH,IN,KT,WV",
  xlab = "Percent Below Poverty",
  col = "darkgreen" 
)

```

###Task 4
```{r }

tm_shape(four_states) +
  tm_polygons("poverty_pct", 
              style = "quantile",
              palette = "greens") +
  tm_layout(frame = FALSE)

```

###Task 5
```{r}
#projecting the data
fs_projected <- st_transform(four_states, "ESRI:102010")

#queen contiguity
nb <-poly2nb(fs_projected, queen = TRUE)
nb

```

####Task5.1
```{r}
#standardize 
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
lw
```

####Task5.2
```{r histogram}

neighbors <- attr(lw$weights, "comp")$d

hist(
  neighbors,
  main = "Number of Neighbors Histogram",
  xlab = "Number of Neighbors",
  col = "lightgreen"
)

```

####Task5.3
```{r average}

mean(neighbors)
#[1] 5.571831

```

####Task5.4
```{r Moran Plot}

moran.plot(fs_projected$poverty_pct, lw, zero.policy = TRUE)

```

###Task 6
```{r}
#for x,y coordinates
coordinates <- st_coordinates(st_centroid(fs_projected))

#distance based neighbors
nb_idw <- dnearneigh(coordinates, 0, 300000)

#IDW
dist_list <- nbdists(nb_idw, coordinates) 
idw_weights <- lapply(dist_list, function(x) 1 / x)

```

####Task6.1
```{r row standardize}

lw_idw <- nb2listw(nb_idw, glist = idw_weights, style = "W", zero.policy = TRUE)

```

####Task6.2
```{r histogram}

neigh_idw <- attr(lw_idw$weights, "comp")$d

hist(
  neigh_idw,
  main = "Number of IDW Neighbors",
  xlab = "Number of Neighbors",
  col = "forestgreen"
)

```

####Task6.3
```{r average}

mean(neigh_idw)
#[1] 0.001266765
```

####Task6.4
```{r Moran Plot}

moran.plot(fs_projected$poverty_pct, lw_idw, zero.policy = TRUE)

```

#Bonus?
```{r}
#to get the data
#put it into a dataframe
#quadrants
#plot

#data behind the plot
fs_moran <- moran.plot(
  fs_projected$poverty_pct,
  lw,
  zero.policy = TRUE,
  plot = FALSE
)

#dataframe
df <- data.frame(
  x = fs_moran$x,
  lag = fs_moran$wx
)
#outliers separate because I couldn't get it to work with the above code
if("is.inf" %in% names(fs_moran)){
  df$significant <- fs_moran$is_inf
}

#vertical line
mean_x <- mean(df$x)

#horizontal line
mean_lag <- mean(df$lag)

#quadrants
df <- df %>%
  mutate(
    quadrant = case_when(
      x >= mean_x & lag >= mean_lag ~ "HH", #high value clusters
      x >= mean_x & lag < mean_lag ~ "HL", #outliers
      x < mean_x & lag < mean_lag ~ "LL", #low value clusters
      x < mean_x & lag >= mean_lag ~ "LH") #outliers
  )

#could not figure out the plot

```

#QUESTIONs
###Q1
####### Morans I is calculated by creating a weight matrix (contiguity or IDW) that is the spatial aspect. The variables are averaged and the the deviations are found to figure out how far from average they are. It compares values and calculates if those values are near each other to measure the spatial autocorrelation.

###Q2
###### A spatially-lagged variable is the summary of a variable from neighboring locations. The values are confined to a single varible using the spatial weights matrix. It provides insight to spatial dependence and whether or not it has an influence on other locations. 

###Q3
###### Contiguity defines neighbors based off of the borders touching while the IDW defines neighbors by distance. Defining W changes the way that spatial patterns are detected. The contiguity Moran plot is closer to the fit line and the outliers are more obvious. The IDW Moran plot is more spread is more spread out and has less outliers. This represents the contiguity method and the more localized patterns vs the IDW wich gives a more accurate picture because of the weight is closer to the counties. 

###Q4
###### A H-L quadrant is a high-low quandrant. High values surrounded by neighboring locations with low values. It's useful because it can identify spatial clustering and provide understanding for patterns for locations and when more data or a different W might be needed to fully complete data analysis. 



